---
title: "ramsey_mercury"
author: "Denver Link"
date: "2023-10-19"
output: html_document
editor_options: 
  chunk_output_type: console
---

#library
```{r}
library(tidyverse)
library(lubridate)
```

#data
```{r}
hg <- read_csv("Scripts/data/ramsey_county_hg.csv")

#take a look at the data
#a row is a mercury value of a fish at a length in a lake
#some surveys aggregated fish of similar length - the number of fish used in the sample is noted by NOFISH
glimpse(hg)
```

#filtering
```{r}
hg <- hg %>% 
  mutate(date = mdy(DATECOL2)) %>% 
  mutate(DOWID = str_replace_all(DOWID, "-", "")) %>% 
  mutate(species = case_when(SPEC == "BLC" ~ "black_crappie",
                             SPEC == "BLG" ~ "bluegill",
                             SPEC == "CAP" ~ "carp",
                             SPEC == "LMB" ~ "largemouth_bass",
                             SPEC == "NOP" ~ "northern_pike",
                             SPEC == "SMB" ~ "smallmouth_bass",
                             SPEC == "WAE" ~ "walleye",
                             SPEC == "WTS" ~ "white_sucker",
                             SPEC == "YEP" ~ "yellow_perch",
                             TRUE ~ SPEC)) %>% 
  filter(!is.na(LGTHCM)) %>% 
  filter(!is.na(HGPPM)) %>% 
  mutate(lake.zm = case_when(is.na(invasion_status) ~ "Uninvaded",
                             invasion_status %in% c("Uninvaded", "ZM") ~ "Zebra Mussel"))
```

#data exploration
```{r}
glimpse(hg)

#lakes in the data
hg %>% 
  group_by(WATERWAY) %>% 
  count() %>% 
  print(n = nrow(.))

#fish from the Missississippi River?
hg %>% 
  filter(WATERWAY == "MISSISSIPPI R.") %>% 
  glimpse()

#coverage of mercury samples from zebra mussel lakes with uninvaded and invaded time periods in all lakes and species
hg %>% 
  filter(!is.na(invasion_status)) %>% 
  ggplot() +
  geom_point(aes(x = LGTHCM, y = HGPPM, color = invasion_status)) +
  geom_smooth(aes(x = LGTHCM, y = HGPPM, color = invasion_status)) +
  facet_grid(SPEC~WATERWAY, scales = "free") +
  scale_color_manual(values = c("lightblue", "salmon"))

#looking at just wae and nop samples
hg %>% 
  filter(SPEC %in% c("WAE", "NOP")) %>% 
  filter(!is.na(invasion_status)) %>% 
  ggplot() +
  geom_point(aes(x = LGTHCM, y = HGPPM, color = invasion_status)) +
  geom_smooth(aes(x = LGTHCM, y = HGPPM, color = invasion_status)) +
  facet_grid(species~WATERWAY, scales = "free") +
  scale_color_manual(values = c("lightblue", "salmon")) +
  theme(legend.position = element_blank())
ggsave("zm_lakes_wae_nop.png", width = 7, height = 5)

hg %>% 
  filter(SPEC %in% c("WAE", "NOP")) %>% 
  filter(!is.na(invasion_status)) %>%
  ggplot() +
  geom_point(aes(x = LGTHCM, y = HGPPM, color = invasion_status)) +
  geom_smooth(aes(x = LGTHCM, y = HGPPM, color = invasion_status)) +
  facet_wrap(species~WATERWAY, scales = "free") +
  scale_color_manual(values = c("lightblue", "salmon"))

#pulling out lakes with sufficient data from zm lakes
hg %>% 
  filter((WATERWAY == "JOHANNA" & SPEC %in% c("NOP", "YEP", "WAE")) |
           WATERWAY == "BALD EAGLE" & SPEC %in% c("NOP", "WAE")|
           WATERWAY == "EAST VADNAIS" & SPEC %in% c("WAE", "NOP")|
           WATERWAY == "LONG" & SPEC %in% c("CAP", "NOP", "WAE") |
           WATERWAY == "MCCARRON" & SPEC %in% c("NOP", "LMB")|
           WATERWAY == "OWASSO" & SPEC %in% c("NOP", "WAE", "YEP", "LMB", "BLG")|
           WATERWAY == "PLEASANT" & SPEC %in% c("NOP", "WAE", "SMB", "WHB")|
           WATERWAY == "WHITE BEAR" & SPEC %in% c("NOP", "WAE")) %>% 
  ggplot() +
  geom_point(aes(x = LGTHCM, y = HGPPM, color = invasion_status)) +
  geom_smooth(aes(x = LGTHCM, y = HGPPM, color = invasion_status)) +
  facet_grid(species~WATERWAY, scales = "free") +
  scale_color_manual(values = c("lightblue", "salmon"))

#lakes that have pre and post data
hg %>% 
  filter((WATERWAY == "JOHANNA" & SPEC %in% c("NOP")) |
           WATERWAY == "BALD EAGLE" & SPEC %in% c("WAE")|
           WATERWAY == "EAST VADNAIS" & SPEC %in% c("WAE")|
           WATERWAY == "LONG" & SPEC %in% c("NOP") |
           WATERWAY == "PLEASANT" & SPEC %in% c("WAE", "WHB") |
           WATERWAY == "WHITE BEAR" & SPEC %in% c("WAE")) %>% 
  ggplot() +
  geom_point(aes(x = LGTHCM, y = HGPPM, color = invasion_status)) +
  geom_smooth(aes(x = LGTHCM, y = HGPPM, color = invasion_status)) +
  facet_grid(species~WATERWAY, scales = "free") +
  scale_color_manual(values = c("lightblue", "salmon"))
ggsave("zm_lakes_pre_post_sufficient.png", width = 7, height = 5)

hg %>% 
  filter(!is.na(invasion_status)) %>% 
  ggplot() +
  geom_boxplot(aes(x = invasion_status, y = HGPPM)) +
  facet_grid(species~WATERWAY, scales = "free")

#how do invaded and uninvaded lakes compare? and are the sampled fish of similar size?
hg %>% 
  ggplot() +
  geom_boxplot(aes(y = log(HGPPM), x = invasion_status)) 

hg %>% 
  ggplot() +
  geom_boxplot(aes(y = LGTHCM, x = invasion_status)) +
  facet_wrap(~SPEC, scales = "free")

#how do they look over time?
hg %>% 
  ggplot() +
  geom_smooth(aes(YEARCOLL, HGPPM, color = invasion_status)) +
  geom_jitter(aes(YEARCOLL, HGPPM, color = invasion_status)) +
  scale_color_manual(values = c("lightblue", "salmon", "black"))

hg %>% 
  ggplot() +
  geom_smooth(aes(YEARCOLL, HGPPM, color = invasion_status)) +
  geom_point(aes(YEARCOLL, HGPPM, color = invasion_status)) +
  scale_color_manual(values = c("lightblue", "salmon", "black"))

#are there any non-invaded samples from 2021?
hg %>% 
  filter(YEARCOLL >= 2021) %>% 
  group_by(WATERWAY, invasion_status,YEARCOLL) %>%
  count()
  print(n = nrow(.))
#only long, johana, and pleasant were sampled in 2021 (thus all after invasion)

#uninvaded lakes samples by speices 
hg %>% 
  filter(is.na(invasion_status)) %>% 
  ggplot() + 
  geom_smooth(aes(LGTHCM, HGPPM)) +
  geom_point(aes(LGTHCM, HGPPM)) +
  facet_grid(species~WATERWAY, scales = "free")

hg %>% 
  filter(is.na(invasion_status)) %>% 
  filter(WATERWAY %in% c("GERVAIS", "NORTH STAR STEEL", "PHALEN", "TURTLE", "SILVER")) %>% 
  ggplot() + 
  geom_smooth(aes(YEARCOLL, HGPPM)) +
  geom_point(aes(YEARCOLL, HGPPM)) +
  facet_grid(species~WATERWAY, scales = "free")

hg %>% 
  filter(is.na(invasion_status)) %>% 
  filter(SPEC %in% c("NOP", "WAE", "CAP", "YEP")) %>% 
  filter(WATERWAY %in% c("GERVAIS", "NORTH STAR STEEL", "PHALEN", "TURTLE", "SILVER")) %>% 
  ggplot() + 
  geom_smooth(aes(LGTHCM, HGPPM)) +
  geom_point(aes(LGTHCM, HGPPM)) +
  facet_grid(species~WATERWAY, scales = "free")
ggsave("best_reference.png", width = 7, height = 5)

#temporal coverage?
hg %>% 
  filter(SPEC %in% c("NOP", "WAE")) %>% 
  ggplot() +
  geom_smooth(aes(YEARCOLL, HGPPM, color = lake.zm)) +
  geom_point(aes(YEARCOLL, HGPPM, color = lake.zm), alpha = .3) +
  facet_wrap(~species, scales = "free") +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("lightblue", "salmon"))
ggsave("temporal_invaded_uninvaded.png", width = 7, height = 5)

#should we concider LOD?
hg %>% 
  filter(!is.na(HGCODE)) %>% 
  group_by(HGCODE, SPEC, WATERWAY) %>% 
  count() %>% 
  print(n = nrow(.))
```

Quick thoughts:
1. it would be great to see if we caught get 2022 and maybe even 2023 data
2. There is a large lack of data for most species/lakes
3. the best ZM pre/post systems are walleye from bald eagle, east vadnais, and pleasant; northern pike from johanna. 
4. lakes with good uninvaded data without post data are northern pike in bald eagle, owasso, mccarron, and white bea; walleye in owasso and white bear; perch in johanna and owasso; largemouth bass in owasso
4. in the lakes with good pre and post data there seems to be a general trend as seen in other lakes
5. The only lakes sampled post 2021 were zebra mussel invaded... might be a problem for BACI
6. several uninvaded lakes have good northern pike data but is sparse otherwise
7. lakes with best uninvaded data: Gervais, norther star steel, phalen, turtle, silver
